---
author: "Chris Reudenbach"
title: "OTB Wrapper"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  html_document: 
    theme: united
    toc: yes
  rmarkdown: default
  pdf_document:
    latex_engine: xelatex
    toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{OTB Wrapper}
  %\VignetteEncoding{UTF-8}{inputenc}\
  %\VignetteEngine{knitr::knitr}
---


# A typical usecase for the Orfeo Toolbox wrapper
link2GI supports the use of the Orfeo Toolbox with a listbased simple wrapper function. Actually two functions parse the modules and functions syntax dumps and generate a command list that is easy to modify with the necessary arguments.

Usually you have to get the module list first:


```{r, eval=FALSE}
# link to the installed OTB 
otblink<-link2GI::linkOTB()


# get the list of modules from the linked version
algo<-parseOTBAlgorithms(gili = otblink)
```

Based on the modules of the current version of `OTB` you can then choose the module(s) you want to use.


```{r, eval=FALSE}
## for the example we use the edge detection, 
algoKeyword<- "EdgeExtraction"

## extract the command list for the choosen algorithm 
cmd<-parseOTBFunction(algo = algoKeyword, gili = otblink)

## print the current command
print(cmd)
```

Admittedly this is a very straightforward and preliminary approach. Nevertheless it provids you a valid list of all `OTB` API calls that can easily manipulated for your needs. The following working example will give you an idea how to use it.


```{r eval=FALSE}
require(link2GI)
require(raster)
require(listviewer)

otblink<-link2GI::linkOTB()
 projRootDir<-tempdir()
 
data('rgb', package = 'link2GI')  
raster::plotRGB(rgb)
r<-raster::writeRaster(rgb, 
              filename=file.path(projRootDir,"test.tif"),
              format="GTiff", overwrite=TRUE)
## for the example we use the edge detection, 
algoKeyword<- "EdgeExtraction"

## extract the command list for the choosen algorithm 
cmd<-parseOTBFunction(algo = algoKeyword, gili = otblink)

## get help using the convenient listviewer
listviewer::jsonedit(cmd$help)

## define the mandantory arguments all other will be default
cmd$input  <- file.path(projRootDir,"test.tif")
cmd$filter <- "touzi"
cmd$channel <- 2
cmd$out <- file.path(projRootDir,paste0("out",cmd$filter,".tif"))

## run algorithm
retStack<-runOTB(cmd,gili = otblink)

## plot filter raster on the green channel
plot(retStack)








```

