---
title: "OTB Wrapper in link2GI"
author: "Chris Reudenbach"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OTB Wrapper in link2GI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

The Orfeo Toolbox (OTB) provides a collection of command-line applications for remote sensing and image processing. `link2GI` offers an R wrapper around these applications with a focus on **reproducibility, transparency, and robustness across OTB versions**.

Historically, `link2GI` exposed OTB applications through a heuristic list-based wrapper ("legacy API"). This is retained for backward compatibility. In addition, `link2GI` now provides a **new introspection-based API** that derives parameter metadata from OTB’s own CLI help output and avoids implicit assumptions.

## Two APIs: legacy vs. introspection

### Legacy API (`parseOTBFunction()` / `runOTB()`)

The legacy API parses the output of:

```
otbcli <algorithm> -help
```

and converts it into a modifiable R list representing the command-line call.

Characteristics:

* simple and permissive
* minimal upfront constraints
* convenient for older scripts

Limitations:

* required parameters and defaults are inferred heuristically
* sensitive to changes in OTB help formatting
* less suitable for strict, version-stable workflows

### Introspection-based API (recommended)

The new API treats the OTB CLI help output as the **single source of truth** for parameters and their status (mandatory vs. optional) and supports a strict "write-to-disk" workflow.

Core functions:

* `otb_capabilities()`: retrieve CLI help text (source of truth)
* `otb_args_spec()`: normalized parameter spec table parsed from `-help`
* `otb_build_cmd()`: create a `runOTB()`-compatible command template (valid keys only)
* `otb_set_out()`: set file-based outputs safely (including `[pixel]` outputs)

---

## Minimal workflow

```{r eval=FALSE}
library(link2GI)
otb <- link2GI::linkOTB(searchLocation="~/apps/otb911/")
cmd <- link2GI::otb_build_cmd("DimensionalityReduction", otb, include_optional="defaults", require_output=TRUE)
cmd[["in"]] <- "in.tif" 
cmd[["method"]] <- "pca"
cmd[["nbcomp"]] <- "3"
cmd <- link2GI::otb_set_out(cmd, otb, key="out", path="out.tif")
str(cmd)
```


## Typical interactive workflow

The following example demonstrates a complete workflow:

1. link an installed OTB
2. read and display the raw `-help` output (source of truth)
3. inspect the parsed parameter table (`spec`)
4. build a command template from *valid keys*
5. set input and explicit on-disk output
6. run via `runOTB()` and plot the result

### Example: `DimensionalityReduction` (PCA) — fully transparent

This example avoids “guessing” parameter names. The only manual step is choosing the correct parameter key for “number of PCA components” by reading the `spec` table printed from your local OTB build.

```{r eval=FALSE}
library(link2GI)
library(terra)
library(mlr3spatial)

# 0) Link OTB
otb <- link2GI::linkOTB(searchLocation="~/apps/otb911/")

# 1) Choose algorithm
algo <- "DimensionalityReduction"

# 2) Read the OTB help text (source of truth)
caps <- link2GI::otb_capabilities(algo = algo, gili = otb, include_param_help = FALSE)
cat(paste(head(caps$text, 60), collapse = "\n"), "\n")

# 3) Get the parsed parameter table (still transparent)
spec <- link2GI::otb_args_spec(algo, otb)
print(spec[, c("key", "class", "mandatory", "default")], row.names = FALSE)

# Build the preliminary command template (only valid keys)
cmd <- link2GI::otb_build_cmd(
  algo,
  otb,
  include_optional = "defaults",
  require_output = TRUE
)



# Identify the PCA component-count key by inspecting 'spec'
# (no fallback list, no heuristics)
pca_related <- spec[grepl("^method\\.pca\\.|^nbcomp$|outdim|comp", spec$key), ]
print(pca_related[, c("key", "mandatory", "default", "desc")], row.names = FALSE)

# Pick the correct key from the printed table 
# (example for OTB 9.1.1: "nbcomp") and fill it with your value)
cmd[["nbcomp"]]   <- "9"



# Input + output (explicit on-disk path, not tempdir)
infile <- system.file("extdata", "leipzig_raster.tif", package = "mlr3spatial")
out_dir <- file.path(getwd(), "otb_tutorial_out")
dir.create(out_dir, recursive = TRUE, showWarnings = T)
out_pca <- file.path(out_dir, "leipzig_pca.tif")[1]

cmd[["in"]] <- infile
cmd[["out"]] <- out_pca


# PCA settings (explicit)
cmd[["method"]] <- "pca"


# Show the exact CLI that will be executed
cat(link2GI::runOTB(cmd, otb, retCommand = TRUE, retRaster = TRUE), "\n")
str(cmd)

# Run + plot
pca_rast <- link2GI::runOTB(cmd, otb, quiet = FALSE)
file.exists(cmd[["out"]])
```



## Legacy workflow (supported but not recommended)

The legacy workflow builds a command list by parsing `-help` output and then executes it.

```r
library(link2GI)
library(terra)

otb <- link2GI::linkOTB()

algo <- "EdgeExtraction"
cmd  <- parseOTBFunction(algo = algo, gili = otb)

# legacy uses input_in / input_il mappings

cmd$input_in <- system.file("ex/elev.tif", package = "terra")

# set required parameters (example values; check cmd$help)

cmd$filter  <- "touzi"
cmd$channel <- 1

# explicit output path (recommended even in legacy)

cmd$out <- tempfile(fileext = ".tif")

res <- runOTB(cmd, gili = otb, quiet = FALSE)
plot(res)
```


## What changed and why

Older versions of the wrapper tried to be convenient by guessing defaults and required parameters. That is fragile across OTB releases.

The introspection-based API:

* trusts OTB metadata instead of heuristics
* separates capability discovery from execution
* enforces explicit file outputs (reproducible artifacts)
* makes failures explicit rather than silent

## Compatibility notes

* The introspection-based API is intended for OTB installations where the CLI `-help` output provides a stable, parseable `Parameters:` section.
* If your OTB version does not expose a parseable parameter block, the introspection parser may return an empty spec; in that case the legacy API may still work (best-effort).

## Migration hints

* Existing scripts that use `parseOTBFunction()` and `runOTB()` should continue to work.
* For new projects, prefer `otb_capabilities()` + `otb_args_spec()` + `otb_build_cmd()` + `otb_set_out()` and then execute via `runOTB()`.


